---
title: "ASM - Module 4 Solutions"
author: "Michal Siwek"
date: "Monday, November 16, 2015"
output: pdf_document
---

## Task 3
Collecting data and fitting the model:
```{r}
file <- "realest.txt"
data <- read.table(file, header = T)
fit <- lm(Price ~ ., data = data)
```
### a) Diagnostic plots
```{r}
par(mfrow = c(2, 2))
plot(fit, which = 1:4)
```

**Residual plot** shows heteroscedasticity - the variance is higher for lower fitted values, and decreases as the fitted values increase. **Normal QQ plot** indicates the distribution of residuals has lighter tails than the normal distribution. **Cook's distance plot** indicates some outlying influential observations, especially observation 8, but also 11 and 6.

### b) Outliers
Based on the heuristic rule we conclude that there are three outliers:
```{r}
res <- rstudent(fit)
res[abs(res) > 2]
```
### c) Influential observations
Observations having highest Cook's distance:
```{r}
cook <- cooks.distance(fit)
cook[order(cook, decreasing = T)][1:3]
```
Influential observation according to the heuristinc rule for the values of leverage:
```{r}
hat <- hatvalues(fit)
threshold <- 2 * sum(hat) / nrow(data)
hat[hat > threshold]
```
## Task 5
Collecting data:
```{r}
file <- "strongx.txt"
data <- read.table(file, header = T)
```
### a) LS fit
```{r}
fitLS <- lm(crossx ~ energy, data = data)
```
### b) WLS fit
```{r}
fitWLS <- lm(crossx ~ energy, weights = sd^-2, data = data)
```
### c) Models comparison
Models output:
```{r}
summary(fitLS)
summary(fitWLS)
```
Diagnostic plots:
```{r}
par(mfrow = c(2, 2))
plot(fitLS)
plot(fitWLS)
```

We can see that the **WLS model (green in the plot below)** has less scattered residuals for small `energy` values (these are actualy small values of the inverse of the energy) than the **LS model (red)**.
```{r}
plot(data$energy, fitLS$residuals, ylim = c(-15, 35), col = 2,
     main = "Residuals of LS and WLS models")
points(data$energy, fitWLS$residuals, col = 3)
abline(h = 0, lty = 2)
```

We can explain this by pointing to the fact that observations with high `energy` have also high values of estimated standard deviation (see the plot below). It means that these observations are multiplied by relativaly smaller weights and *so the observations with low `energy` values have stronger impact on the regression line leading to a better fit for them*.
```{r}
plot(data$energy, data$sd,
     main = "Estimated standard deviation")
```

### d) Modification - WLS2 model
The residual plot shows nonlinearity: a convex depenence of actual `crossx` values not caught by the regression on `energy`. Thus we could **include square values of `energy`** to account for this dependence:
```{r}
fitWLS2 <- lm(crossx ~ energy + I(energy^2), weights = sd^-2, data = data)
summary(fitWLS2)
```
We see the squared term is significant, the adjusted $R^2$ is $99\%$, so much better than the previous WLS model ($93\%$).

The diagnostic plots:
```{r}
par(mfrow = c(2, 2))
plot(fitWLS2)
```

The plots show that there are still many problems with the fit.

### e) Drawing fitted curves
```{r}
par(mfrow = c (1, 1))
plot(data$energy, data$crossx,
     main = "Fitted curves of all three models")
lines(data$energy, fitLS$fitted.values, col = 2)
lines(data$energy, fitWLS$fitted.values, col = 3)
lines(data$energy, fitWLS2$fitted.values, col = 4)
legend(x = 0.06, y = 350,
       legend=c("LS", "WLS", "WLS2"),
       lty = 1, col = 2:4)
```
